---
title: "Bayesian Network Analysis"
subtitle: "Cell-specific sn-RNAseq"
author:  |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
params:
  results_dir: "/pastel/projects/speakeasy_dlpfc/BN/edges_restricted/" # Will create a folder named by data_prefix here
  data_prefix: "app07_sn_mic12_mic13_ast10_bonf001_runt500_impute10PCs_restrict_tangles2amyloid" # Prefix for the results
  SEnet_dir: "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/results/all_cells/" # SE results folder. Expect files named "lv3_moduleEigengenes.Rdata" and "geneBycluster.txt"
  pval_cutoff: 0.01 # Bonf
  runt: 500
  covariates: ["cogng_demog_slope","tangles_sqrt","amyloid_sqrt"]
  cellFrac_file: "/pastel/resources/20220203_snRNAseq_AMPAD/updated_annotations/subcellprop_snRNAseq.txt"
  cells2include: ["Mic.13","Ast.10","Mic.12"]
  use_eigengene: FALSE # If FALSE, use AverageExpression
  BN_run_dir: "/pastel/resources/bayesian_networks/CINDERellA_new/" # no need to edit
  resources_dir: "/pastel/resources/" # no need to edit
  restrict_edges: TRUE
  restrict_edges_from: ["tangles"] # Should map 1:1 with restrict_edges_to
  restrict_edges_to: ["amyloid"] # Should map 1:1 with restrict_edges_from
output: 
 rmarkdown::html_document:
   theme: united
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(kableExtra)
library(reshape2)
library(dplyr)
library(network)
library(ggraph)
library(GGally)
library(sna)
library(ggplot2)
library(ggnetwork)
library(ggraph)
library(tidygraph)
library(ggsci)
library(data.table)
library(reshape2)
library(tidymodels)
library(furrr)
library(R.matlab)
library(BiocParallel)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(gtools)
library(graphlayouts)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE)
```

```{css zoom-lib-src, echo = FALSE, eval = F}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r Helper Functions, echo=FALSE}
# Enrichment
source("/pastel/Github_scripts/SpeakEasy_net/downstream_speakeasy/speakeasy_downstream_lib.R") # Fisher test
source("/pastel/Github_scripts/OtherProjects/useful_scripts/common_lib.R")
```

## Run parameters 

```{r}
# Prefix for the results
data_prefix = params$data_prefix

# Modules to include in the BN
pval_cutoff = params$pval_cutoff # P-value cutoff for module-trait associations
runt = params$runt # Runtime for CINDERellA

# Add covariates to the module avg matrix (avoid adding more than 3, since traits tend to correlate with each other)
covariates = params$covariates

# Folder with SE module eigengenes
net_dir = params$SEnet_dir

cellFrac_file = params$cellFrac_file # File with cell fractions
cells2include = params$cells2include # Cells to include in the BN

use_eigengene = params$use_eigengene # If FALSE, use AverageExpression
restrict_edges = params$restrict_edges # If TRUE, tangles upstream of amyloid, and cog. decline upstream tangles/amyloid will not be allowed
restrict_edges_from = params$restrict_edges_from
restrict_edges_to = params$restrict_edges_to

BN_run_dir = params$BN_run_dir
resources_dir = params$resources_dir
bn_results_dir = paste0(params$results_dir,"/",data_prefix,"/")
BN_output_dir = paste0(bn_results_dir,data_prefix,"_res")
BN_run_data = paste0(bn_results_dir,data_prefix,"_exp.txt")

print(params)
```

## BN input/parameters 

`no need to edit`

CINDERellA has two inputs: exp.txt and output_folder

Here we create the exp.txt by merging gene/moddule data with traits.

```{r message=T, eval=T}
message("Creating folders...")

system(paste0("mkdir -p ",bn_results_dir))
system(paste0("mkdir -p ",BN_output_dir))

save(params, file = paste0(bn_results_dir,"params.rds"))
```

## Speakeasy input/parameters 

### 1. Phenotypes

Just getting the phenotype matrix and creating some custom phenotypes. This will be filtered later to match samples from the networks.

```{r, message=T, eval=T}
message("Preparing phenotypes...")

load("/pastel/projects/spatial_t/pseudo_bulk/phenotypes.RData") # phenotypes that match with the previous regression analysis 
# load("/pastel/projects/spatial_t/pseudo_bulk/phenotypes_includeZeros.RData") # phenotypes
phenotype_dt = phenotypes

save(phenotype_dt, file = paste0(bn_results_dir,"phenotypes.RData"))

message("Saved: ", paste0(bn_results_dir,"phenotypes.RData"))
```

### 2. Assign cell-type module annotations

```{r message=T, eval=T}
message("Assigning cell-type module annotations...")

library(purrr)
get_modCellType <- function(gene_net){
  path_to_files = "/pastel/resources/gene_lists/celltype_markers/"
  geneList_file_names = c("m109_mostafavi.txt",
                          "GWAS_AD_Bel2022.txt")
  genes_universe = gene_net$ensembl
  cluster_lables = gene_net$module
  geneAnnotation_lists = parseGeneLists(path_to_files, geneList_file_names, genes_universe)
  gene_cluster_enrich = moduleEnrich(genes_universe, cluster_lables, geneAnnotation_lists)
  return(gene_cluster_enrich)
}

# Gene-module assignments
gene_net = read.table(file = paste0(net_dir, "geneBycluster.txt"), sep = "\t", header = T)

gene_net$module = paste0("ME",gene_net$cluster_lv3)
gene_net = gene_net %>% group_by(module) %>% mutate(module_size = n())
gene_net = gene_net %>% filter(module_size >= 30)

# Get enrichments using gene lists
module_gene_celltype = get_modCellType(gene_net)

module_gene_celltype_top = module_gene_celltype[module_gene_celltype$p_bonferroni < 0.05,] %>% 
  group_by(module_id) %>% 
  mutate(module_id, annotation = paste0(unique(gsub("(.*?) (.*)","\\1",gene_list)), collapse = ";")) %>%
  select(module_id,annotation) %>% 
  distinct()

gene_net_annot = gene_net %>% 
  left_join(module_gene_celltype_top, by = c("module"="module_id"))

gene_net_metadata = gene_net_annot %>% 
  dplyr::select(module, module_size, annotation) %>%
  distinct() %>% arrange(-module_size)

save(gene_net_metadata, file = paste0(bn_results_dir,"gene_net_metadata.RData"))

message("Saved: ", paste0(bn_results_dir,"gene_net_metadata.RData"))
```

### 3. Load Eigengenes

```{r message=T, eval=T}
message("Loading SE modules...")

load(paste0(bn_results_dir,"gene_net_metadata.RData"))
load(paste0(bn_results_dir,"phenotypes.RData"))

net_modules = read.table(file = paste0(net_dir, "geneBycluster.txt"), sep = "\t", header = T)
modules_size = as.data.frame( table(net_modules$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")
modules_size$module = paste0("ME",modules_size$module)

# Eigengenes and average expression 
load(paste0(net_dir, "lv3_moduleEigengenes.Rdata"))
if(use_eigengene){
  message("Using eigengenes")
  mod_eigengene = lv3_moduleEigengenes$eigengenes  
}else{
  message("Using average expression")
  mod_eigengene = lv3_moduleEigengenes$averageExpr
  colnames(mod_eigengene) = gsub("^AE","ME",colnames(mod_eigengene))
}

mod_eigengene$projID = gsub("(.*)_(.*)", "\\2", rownames(mod_eigengene)) # get the projid to match with phenotype data 
# all(mod_eigengene$projID %in% phenotype_dt$projid) # Must be TRUE
rownames(mod_eigengene) = mod_eigengene$projID
mod_eigengene$projID = NULL

to_show = modules_size$module[modules_size$n_nodes >= 30]
mod_eigengene = mod_eigengene[,to_show]

# Add a global average module 
mod_eigengene$ME0_global_avg = rowMeans(mod_eigengene)

save(mod_eigengene, modules_size, file = paste0(bn_results_dir,"mod_eigengene.RData"))
message("Saved: ", paste0(bn_results_dir,"mod_eigengene.RData"))
```

### 4. Select Modules based on (pre-computed) association results

```{r message=T, eval=T}
message("Selecting modules based on association results...")

load(paste0(bn_results_dir,"gene_net_metadata.RData"))
load(paste0(bn_results_dir,"phenotypes.RData"))
load(paste0(bn_results_dir,"mod_eigengene.RData"))

work_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/figures4paper/v2_mar2024/"
assoc_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/sn_dlpfc/2nd_pass/eigen_reports/save_lr_adjcov/"

# Non-modules to be removed (< 30 nodes)
emods2remove = read.table(paste0(work_dir, "non_modules.txt"), header = T, stringsAsFactors = F)

# Load association results 
load(paste0(assoc_dir, "all_res_test_stats_SN.Rdata"))
all_stats$module_temp <- paste0(all_stats$network, "_M", all_stats$module)
all_stats$module2 <- gsub("MAE", "M", all_stats$module_temp)
all_stats$module_temp = NULL 

# Remove non-modules 
all_stats_filt <- all_stats[! all_stats$module2 %in% emods2remove$module2, ]
all_stats_filt$network = as.factor(all_stats_filt$network)

pheno_variable = c("cogng_demog_slope","amyloid_sqrt","tangles_sqrt")

all_stats_bonf = all_stats_filt %>% 
  filter(phenotype %in% pheno_variable) %>%
  group_by(phenotype) %>% # Adjust by phenotype separately
  mutate(bonf_p = p.adjust(nom_p, method = "bonferroni")) %>%
  filter(bonf_p <= pval_cutoff) # Bonferroni correction for all tests (modules X phenotypes; 1544 tests)

signifMods = unique(all_stats_bonf$module2)
signifMods = gsub("(.*)_M(.*)","ME\\2_\\1",signifMods)

save(signifMods, file = paste0(bn_results_dir,"mod_Select.RData"))
message("Saved: ", paste0(bn_results_dir,"mod_Select.RData"))
```

### 5. Cell fractions

```{r, message=T, eval=T}
message("Preparing cell fractions...")

cell_prop = read.delim2(cellFrac_file, 
                        header = T, 
                        colClasses = "character", 
                        stringsAsFactors = F, 
                        check.names = F)

cell_prop = cell_prop %>% column_to_rownames("projid")
cell_prop = as.data.frame(t(cell_prop)) %>% mutate_if(is.character, as.numeric)

# Impute missing values for cell_fraction
if (!require("missMDA")) install.packages("missMDA")
cell_prop.imputed = missMDA::imputePCA(cell_prop,ncp=10)
cell_prop = cell_prop.imputed$completeObs
cell_prop[cell_prop < 0] = 0

cell_prop = as.data.frame(t(cell_prop[cells2include,])) %>% na.omit()

save(cell_prop, file = paste0(bn_results_dir,"cell_prop.RData"))
message("Saved: ", paste0(bn_results_dir,"cell_prop.RData"))
```

### 6. Write BN inputs

```{r message=T, eval=T}
message("Writing BN inputs...")

load(paste0(bn_results_dir,"gene_net_metadata.RData")) # gene_net_metadata
load(paste0(bn_results_dir,"phenotypes.RData")) # phenotype_dt
load(paste0(bn_results_dir,"mod_eigengene.RData")) # mod_eigengene
load(paste0(bn_results_dir,"cell_prop.RData")) # cell_prop
load(paste0(bn_results_dir,"mod_Select.RData")) # signifMods

mod_eigengene = mod_eigengene[,signifMods]
# Matching all tables
donor_match_1 = base::intersect(unique(phenotype_dt$projid), rownames(mod_eigengene))
donor_match_2 = base::intersect(donor_match_1, rownames(cell_prop))
donor_match = donor_match_2

phenotype_match = phenotype_dt[match(donor_match,phenotype_dt$projid),]
mod_eigengene_match = mod_eigengene[match(donor_match,rownames(mod_eigengene)),]
cell_prop_match = cell_prop[match(donor_match,rownames(cell_prop)),]
cell_prop_match = as.matrix(cell_prop_match)
gene_net_metadata_match = gene_net_metadata[match(colnames(mod_eigengene_match),gene_net_metadata$module),]

#################################################################################
## 1. exp.txt (row = gene/modules/covariates; column = individuals/samples avg; no rownames/colnames)

dataExp = t(na.omit(cbind( mod_eigengene_match, phenotype_match[,covariates,drop=F], cell_prop_match )))
# dim(dataExp) # 19 402
# Save dataExp to csv
write_csv(as.data.frame(dataExp), file = BN_run_data, col_names = F)

#################################################################################
## 2. allowed_edges.csv (row = from; to; allowed; no rownames/colnames)

# Restrict allowed edges. If 1 then connection is allowed. If 0 then connection is not allowed.
# Format must be a csv table with 3 columns: from, to, allowed. No rownames/colnames, just the positions from dataExp matrix
allowed_edges = matrix(1, nrow = nrow(dataExp), ncol = nrow(dataExp))
rownames(allowed_edges) = rownames(dataExp)
colnames(allowed_edges) = rownames(dataExp)
allowed_edges_m = reshape2::melt(allowed_edges)
colnames(allowed_edges_m) = c("from","to","allowed")

if(restrict_edges){
  for(i in 1:length(restrict_edges_from)){
    match_from = allowed_edges_m$from %in% grep(restrict_edges_from[i],allowed_edges_m$from, value=T)
    match_to = allowed_edges_m$to %in% grep(restrict_edges_to[i],allowed_edges_m$to, value=T)
    allowed_edges_m$allowed[match_from & match_to] = 0
  }
}

# Rename nodes to their respective positions
allowed_edges_m$from = match(allowed_edges_m$from,rownames(dataExp))
allowed_edges_m$to = match(allowed_edges_m$to,rownames(dataExp))

# Save csv
BN_allowed_edges = paste0(bn_results_dir,"allowed_edges.csv")
write_csv(allowed_edges_m, file = BN_allowed_edges, col_names = F)

#################################################################################

message("Modules included: ", ncol(mod_eigengene_match))
message("Covariates: ", ncol(phenotype_match[,covariates,drop=F]))
message("Cell fractions: ", ncol(cell_prop_match))
message("Sample size merged data: ", ncol(dataExp))

save(phenotype_match,mod_eigengene_match,cell_prop_match,gene_net_metadata_match,dataExp,
     file = paste0(bn_results_dir,"BN_input.RData"))
message("Saved: ", paste0(bn_results_dir,"BN_input.RData"))
```

### Run BN

```{r, eval=T}
setwd(BN_run_dir)
cmd_matlab_call = paste0("matlab -nodisplay -nojvm -nosplash -nodesktop -r")

cmd_matlab_param = paste0("runt=",runt,"';",
                          " data='",BN_run_data,"';",
                          " allowed_edges='",BN_allowed_edges,"';",
                          " out_dir='",BN_output_dir,"';",
                          " run('",BN_run_dir,"CINDERellA.m')")
cmd_matlab_run = paste0(cmd_matlab_call, " \"",cmd_matlab_param,"\"")
cat(cmd_matlab_run)
```

```{r, eval=F}
system(cmd_matlab_run)
```

Results are saved here:

```{r}
print(BN_output_dir)
```

## Read results

```{r fig.width=12, fig.height=8, eval=TRUE}
load(file = paste0(bn_results_dir,"BN_input.RData")) 
params_check = readMat(paste0(BN_output_dir,"/Param.mat"))

edgefrq = read_tsv(paste0(BN_output_dir,"/edgefrq.txt"), col_names = c("A","B","freq"), show_col_types = F)

useBestNet = F
if(useBestNet){
  bestNet = read_tsv(paste0(BN_output_dir,"/bmanet.txt"), col_names = c("A","B","freq"), show_col_types = F)
  edgefrq = bestNet %>% mutate(freq = NA) %>%
    left_join(edgefrq, by = c("A","B")) %>%
    mutate(freq = ifelse(is.na(freq.y), freq.x, freq.y)) %>%
    select(-freq.x, -freq.y) %>%
    na.omit()
}

edges_df = na.omit(cbind(edgefrq, rownames(dataExp)[edgefrq$A], rownames(dataExp)[edgefrq$B]))
colnames(edges_df) = c("fromNode", "toNode", "weight", "fromAltName", "toAltName") # match names for Cytoscape input

# write_tsv(edges_df, file = paste0(BN_output_dir,"/edges.tsv"), col_names = T)

# createDT(edges_df %>% arrange(-weight))

edges_filtered = edges_df[abs(edges_df$weight)>0.4, ]
rownames(edges_filtered) = NULL
#edges_filtered %>% arrange(-weight) #%>% filter(fromAltName == "ME3")

nodes_table = data.frame(nodeName = 1:nrow(dataExp), altName = rownames(dataExp)) 
rownames(nodes_table) = NULL
nodes_table = nodes_table %>% left_join(gene_net_metadata_match, by = c("altName"="module"))
nodes_table$nodeAttr = ifelse(grepl("^ME",nodes_table$altName),"Module","Trait")
nodes_table$annotation = ifelse(grepl("^ME",nodes_table$altName),nodes_table$annotation,"Trait")

annot_set = as.character(na.omit(unique(nodes_table$annotation)))
names(annot_set) = pal_d3()(length(annot_set))
annot_set = c(annot_set, grey = NA)
annot_pal = setNames(as.data.frame(annot_set),"annotation") %>% rownames_to_column("color")

nodes_table = nodes_table %>% left_join(annot_pal)

nodes_filtered = nodes_table[nodes_table$nodeName %in% c(edges_filtered$fromNode,edges_filtered$toNode), ]

mypal = c(ast = "#1F77B4", # ast
     end = "#FF7F0E", # end
     ext = "#2CA02C", # ext
     inh = "#D62728", # inh
     mic = "#9467BD", # mic
     opc = "#8C564B", # opc
     oli = "#E377C2", # oli
     cells = "white",
     traits = "grey30"
     )
cell_text_color = setNames(mypal[tolower(gsub("(.*)\\.(.*)","\\1",cells2include))],cells2include)
nodes_filtered$text_color = ifelse(nodes_filtered$altName %in% names(cell_text_color), cell_text_color[nodes_filtered$altName], "black")

nodes_filtered$celltype = ifelse(nodes_filtered$nodeAttr == "Module", gsub("(.*)_(.*)","\\2",nodes_filtered$altName), nodes_filtered$altName)
nodes_filtered$nodeAttr[nodes_filtered$celltype %in% cells2include] = "cells"
nodes_filtered$nodeAttr[nodes_filtered$celltype %in% covariates] = "traits"
nodes_filtered$celltype[nodes_filtered$celltype %in% cells2include] = "cells"
nodes_filtered$celltype[nodes_filtered$celltype %in% covariates] = "traits"

edges_filtered_bestLink = edges_filtered[,c("fromAltName","toAltName","weight")] %>% 
  group_by(fromAltName,toAltName) %>%
  mutate(link = paste(sort(c(fromAltName,toAltName)),collapse = "_")) %>% 
  group_by(link) %>% slice_max(order_by = weight, n = 1) %>% distinct() 

library(igraph)
library(ggraph)
library(graphlayouts)
library(ggforce)

# Get the graph object
g = igraph::graph_from_data_frame(d = edges_filtered_bestLink, 
                          vertices = nodes_filtered[,c("altName","nodeAttr","module_size","annotation","color","text_color","celltype")], 
                          directed = T)

# this function is borrowed from the ambient package
normalise <- function(x, from = range(x), to = c(0, 1)) {
  x <- (x - from[1]) / (from[2] - from[1])
  if (!identical(to, c(0, 1))) {
    x <- x * (to[2] - to[1]) + to[1]
  }
  x
}

# map to the range you want
#V(g)$degree <- degree(g, mode = "in")
#V(g)$degree <- normalise(V(g)$degree, to = c(3,11))

V(g)$degree <- normalise(na.replace(V(g)$module_size, mean(V(g)$module_size,na.rm = T)), to = c(10,15))
l <- igraph::layout_as_tree(g, flip.y = T)

# Cairo::CairoPDF(file = "/pastel/Github_scripts/SpeakEasy_dlpfc/figures4paper/v2_mar2024/BN_SN_app07.pdf", width = 10, height = 6.5)
ggraph(g, layout = 'kk') +
  geom_edge_fan(
    aes(end_cap = circle(node2.degree + 4, "pt"),  
        edge_width = weight,                       # Edge width = 3 * freq
        alpha = weight),                             # Edge transparency = freq
    edge_colour = "black",                           # Edge color
    arrow = arrow(angle = 15,                      
                  type = "closed",
                  ends = "last",
                  length = unit(0.15, "inches"))
    ) +
  geom_node_point(aes(size = I(degree),              # Node size overlay
                      #alpha = .5 + degree,           # Node color fade overlay
                      fill = celltype,              # Node color overlay
                      shape = nodeAttr),             # Node shape overlay
                  color = "black",
                  show.legend = T) +
  shadowtext::geom_shadowtext(aes(x, y, label = name), 
                              color = nodes_filtered$text_color, 
                              size = 3.5, 
                              bg.colour = "white") +
  scale_edge_color_continuous(low = "grey66", high = "black") +
  scale_edge_width(range = c(0.1, 0.5)) +
  scale_fill_manual(breaks = names(mypal), values = mypal) +
  scale_shape_manual(values = c(23,22,21)) +
  theme_graph() + 
  theme(legend.position = "right") +
  guides(fill = guide_legend(override.aes = list(size = 5, shape = c(22,22,22,22,22,23,21))),
         shape = "none")
# dev.off()
```


## Session info
```{r}
sessionInfo()
```
