---
title: "SpeakEasy clusters: MF"
subtitle: "Eigengene calculation"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo A. Vialle
 | Shinya Tasaki
 |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> SpeakEasy code from [Gaiteri et al, 2015.](https://www.nature.com/articles/srep16361)

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r Helper Functions, echo = FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r load_packages, message=F, warning=F}
library(tidyverse)
library(janitor)
library(readxl)
library(WGCNA)

net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_net_MF/"
expression_dir = "/pastel/projects/speakeasy_dlpfc/"
dir_mod_02 = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_net_MF/"
```

## Input 
In order to summarize the SE network information we will perform the module eigengene approach. Module eigengene is defined as the first principal component of the expression matrix of the corresponding module.

```{r input}
load(paste0(expression_dir, "exprdata_byregion.Rdata")) 

# Expression data for a single set in the form of a data frame where rows are samples and columns are genes (probes):
expr_matx_t = as.data.frame( t(exprData_MF)) # Residuals of the expression
k_dataset_02 = read.table(paste0(dir_mod_02, "geneBycluster.txt"), header = T, stringsAsFactors = F) # clusters from SpeakEasy
k_dataset_ord = k_dataset_02[match(colnames(expr_matx_t), k_dataset_02$ensembl), ] #order the matrices 
# all(colnames(expr_matx_t) == k_dataset_ord$ensembl) # must be true 

```

## Module eigengenes - Lv3

Showing the module eigengenes in a dataframe, with each column corresponding to one eigengene. 
```{r eigen}
colors = k_dataset_ord$cluster_lv3
lv3_moduleEigengenes = moduleEigengenes(expr_matx_t, colors, verbose = 0)

# save results
save(lv3_moduleEigengenes, file = paste0(net_dir, "lv3_moduleEigengenes.Rdata"))
write.table(lv3_moduleEigengenes$eigengenes, file = paste0(net_dir, "lv3_moduleEigengenes.txt"), sep = "\t", quote = F, row.names = T)

createDT(lv3_moduleEigengenes$eigengenes)
```

## Variance Explained

A dataframe in which each column corresponds to a module, with the component varExplained[PC, module] giving the variance of module explained by the principal component no. PC. The calculation is exact irrespective of the number of computed principal components. At most 10 variance explained values are recorded in this dataframe.

**How much of the 1st PC explains expression of the module?** For example, the Module 01 eigengene explains 15% of the expression of the genes in the module. 

Eigengene of a "Module" with one gene is equal 1. 

```{r var_exp}
write.table(lv3_moduleEigengenes$varExplained, file = paste0(net_dir, "lv3_variancer_exp.txt"), sep = "\t", quote = F, row.names = T)
colnames(lv3_moduleEigengenes$varExplained) = gsub("X", "M", colnames(lv3_moduleEigengenes$varExplained))
createDT(lv3_moduleEigengenes$varExplained)
```

## Average expression - Lv3
A dataframe containing average normalized expression in each module. 
```{r ave}
# The columns are named by the corresponding color with an "AE" prepended, e.g., AEturquoise etc.
colnames(lv3_moduleEigengenes$averageExpr) = gsub("AE", "AE_M", colnames(lv3_moduleEigengenes$averageExpr))
write.table(lv3_moduleEigengenes$averageExpr, file = paste0(net_dir, "lv3_averageExpr.txt"), sep = "\t", quote = F, row.names = T)
createDT(lv3_moduleEigengenes$averageExpr)
```

---

## Session info

```{r}
sessionInfo()
```
