---
title: "SpeakEasy clusters: MF"
subtitle: "Checking the network topology"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo Vialle 
 | 
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r Helper Functions, echo = FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r load_packages, message=F, warning=F}
library(tidyverse)
library(reshape2)
library(ggeasy)

net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_net_MF/"
expression_dir = "/pastel/projects/speakeasy_dlpfc/"
work_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/bulk_dlpfc/"
```

# Top 5 hub genes
Hub genes are the most connected genes in a module so, here we list the top ones by module. 
```{r cor_matrix, message=F}
#############################
# Edges table
#############################

################################## Variables to set
# edge_weight2plot = 0.01 # Filter the edges by this value. This is the fraction of the top edge correlations. 0.1 = top 10%.
edge_weight_1 = 1 # We don't wanna plot the correlations = 1 
correlation_method = "pearson"
################################## Done

# Expression data for a single set:
load(paste0(expression_dir, "exprdata_byregion.Rdata")) 
expr_matx = as.data.frame(exprData_MF) # Residuals of the expression

gene_modules = read.table(paste0(net_dir, "geneBycluster.txt"), header = T, stringsAsFactors = F) # clusters from SpeakEasy
gene_modules_size = gene_modules %>% group_by(cluster_lv3) %>% summarise(n = n()) %>% filter(n >= 30)
n_modules = nrow(gene_modules_size)
  
nodes_df = data.frame()
edges_df = data.frame()

for(i in 1:n_modules){
 # i=1
  mod2plot = gene_modules_size$cluster_lv3[i]
  modSize = gene_modules_size$n[i]
    
  # Select the expression values for the module of interest 
  to_plot = gene_modules$ensembl[gene_modules$cluster_lv3 == mod2plot]
  expr_matx_mod = expr_matx[to_plot, ]
    
  #Let's calculate the correlation matrix
  matx_cor = cor(t(expr_matx_mod), method = correlation_method) # must be gene by gene
  matx_cor_m = melt(matx_cor)
  colnames(matx_cor_m) = c("ensembl1", "ensembl2", "value")
   
  matx_cor_temp = matx_cor_m %>% left_join(gene_modules[, c("ensembl", "symbol")], by=c("ensembl1" = "ensembl"))
  colnames(matx_cor_temp) = c("ensembl1", "ensembl2", "value", "from_node")
    
  # Complete correlation matrix 
  matx_cor_temp2 = matx_cor_temp %>% left_join(gene_modules[, c("ensembl", "symbol")], by=c("ensembl2" = "ensembl"))
  colnames(matx_cor_temp2) = c("ensembl1", "ensembl2", "value", "from_node", "to_node")
    
  # Filter the edges based on correlation (correlation is used as weight here)
  edges_before_filtering = matx_cor_temp2[which(abs(matx_cor_temp2$value) != edge_weight_1), ]
    
  #order_of_edges = order(abs(edges_before_filtering$value), decreasing = T)
  #order_of_edges_filt = order_of_edges[1:round(length(order_of_edges)*edge_weight2plot)]
  #edges_filtered = edges_before_filtering[order_of_edges_filt,]
    
  # edges2keep = which(edges_before_filtering$value >= (median(abs(edges_before_filtering$value)))) # Get the edges > than the median  
  edges2keep = which(edges_before_filtering$value >= quantile(abs(edges_before_filtering$value), 3/4)) # Get the edges on the 3/4 quantiles 
  edges_filtered = edges_before_filtering[edges2keep,]
    
  colnames(edges_filtered) = c("fromNode", "toNode", "weight", "fromAltName", "toAltName") # match names for Cytoscape input
    
  # Remove NAs and duplicated ensembls 
  #edges_filtered = na.omit(edges_filtered)
  #edges_filtered = edges_filtered[! duplicated(edges_filtered), ]
  edges_filtered_df = edges_filtered
  edges_filtered_df$module = mod2plot
  edges_filtered_df$region = "MF"
    
  edges_df = rbind(edges_df, edges_filtered_df) # Filtered edges table with all modules from all cell types 
    
  #############################
  # Nodes table 
  #############################
  nodes_table = gene_modules[ , c("ensembl", "cluster_lv3", "symbol")]
  colnames(nodes_table) = c("nodeName", "nodeAttr", "altName") # Match for Cytoscape input 
  nodes_table = nodes_table[nodes_table$nodeAttr == mod2plot, ] # Only for the module of interest
    
  # Apply same filter to the nodes table
  nodes_filtered = nodes_table[nodes_table$nodeName %in% c(edges_filtered$fromNode,edges_filtered$toNode), ]
    
  # Add number of node connections in the filtered network to get the hub nodes
  nodes_filtered = inner_join(nodes_filtered, 
                              bind_rows(edges_filtered %>% group_by(fromNode) %>% dplyr::count() %>% dplyr::rename(nodeName = fromNode), 
                                        edges_filtered %>% group_by(toNode) %>% dplyr::count() %>% dplyr::rename(nodeName = toNode)) %>% 
                                group_by(nodeName) %>% distinct() %>% tally(wt = n))
    
  hub_genes_list = nodes_filtered[order(nodes_filtered$n, decreasing = T), ]$altName[1:5] # Hubs are the most connected nodes in a module (after filtering the edges)
    
  nodes_filtered_df = nodes_filtered[order(nodes_filtered$n, decreasing = T),]
  nodes_filtered_df$module_size = modSize
  nodes_filtered_df$region = "MF"
  nodes_df = rbind(nodes_df, nodes_filtered_df) # Filtered nodes table with all modules from all cell types 
}

# Top hub genes 
nodes_df_top = nodes_df %>% 
  group_by(region, module_size) %>% 
  arrange(-n) %>% 
  slice_head(n = 5)

# write.table(nodes_df_top, file = paste0(work_dir, "hub_genes_MF.txt"), sep = "\t", quote = F, row.names = F)

createDT(nodes_df_top)

# save(nodes_df, file = paste0(work_dir, "n_connections_MF.Rdata"))
# nodes_df %>% group_by(celltype,module_size) %>% summarise(n_genes = n()) %>% arrange(n_genes)
```

# Plot connections 
We want to check if the network has a scale-free topology. 

**Scale-free:** many nodes with only a few links. A few hubs with large number of links. 

**Random:** most nodes have the same number of links. No highly connected nodes resulting in a Poison distribution. 
```{r, fig.width=5, fig.height=3.5}
ggplot(nodes_df, aes(x=n)) +
  geom_histogram(bins = 100, color="black", fill="white") +
  easy_labs(x = "Number of links (k)", y = "Number of nodes with k links") +
  theme_classic()
```

# Session 
```{r}
sessionInfo()
```

