---
title: "SpeakEasy clusters: MF"
subtitle: "Regressions between the covariates and the eigengenes"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo Vialle 
 | 
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Covariates from the phenotype data | Results adjusted by age, sex and education | Function version March 2023. 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(ggeasy)
library(ggplot2)
library(broom)
library(gplots)
library(ggpubr)
library(tidyverse)
```

```{r Helper Functions}
## Create table 
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}

## Run regression tests
run_module_trait_association <- function(data4linear_reg, # Matrix with module eigengenes (predictor)
                                         phenotype_dt, # Matrix with covariates (outcome + covariates)
                                         pheno_list, # List of phenotypes to be tested (with classes = binomial or gaussian)
                                         covariates = c("age_death","msex", "educ") # List of covariates to be adjusted
                                         ){ 
  
  if (!require("lme4")) install.packages("lme4")
  if (!require("lmerTest")) install.packages("lmerTest")
  if (!require("performance")) install.packages("performance")
  library(lme4)
  library(lmerTest)
  library(performance)
  
  outcome = names(pheno_list)
  outcome.family = pheno_list
  # random_effect = "projid"
  # avg_over_random_effect = T
  
  matrix_rsquared = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average)) #Number of modules
  matrix_pvalue = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  
  for (x in 1:length(pheno_list)){
    for (y in 1:ncol(mod_average)){
      outcome_pheno = outcome[x]
      outcome_type = outcome.family[x]

      dat4test_1 = setNames(as.data.frame(cbind(phenotype_dt[,outcome_pheno],data4linear_reg[,y])), c("outcome","predictor"))
      if(!is.null(covariates)){
        dat4test_2 = phenotype_dt[,covariates,drop=F]
        dat4test = na.omit(cbind(dat4test_1, dat4test_2))
        formula_string = as.formula(paste0("outcome ~ predictor + ", paste(covariates, collapse = " + ")))
        print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y], " + ", paste(covariates, collapse = " + ")))
      }else{
        dat4test = na.omit(dat4test_1)
        formula_string = as.formula(paste0("outcome ~ predictor"))
        print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y]))
      }
      
      if (outcome_type == "gaussian"){
        mod.obj0 = lm(formula_string, dat4test, na.action = "na.exclude")
        matrix_rsquared[x,y] <- summary( mod.obj0 )$adj.r.squared
        matrix_pvalue[x,y] <- summary( mod.obj0 )$coefficients["predictor","Pr(>|t|)"] #To insert pvalues in the heatmap
      }
      if (outcome_type == "binomial"){
        dat4test$outcome = as.factor(dat4test$outcome)
        mod.obj1 = glm(formula_string, dat4test, family = binomial, na.action = "na.exclude")
        matrix_rsquared[x,y] <-  1 - mod.obj1$deviance/mod.obj1$null.deviance # Pseudo r-squared
        matrix_pvalue[x,y] <- coef(summary(mod.obj1))["predictor",'Pr(>|z|)']
      }
    }
  }
  
  rownames(matrix_rsquared) = names(pheno_list)
  rownames(matrix_pvalue) = names(pheno_list)
  colnames(matrix_rsquared) = colnames(data4linear_reg)
  colnames(matrix_pvalue) = colnames(data4linear_reg)
  
  matrix_pvalue_df = setNames(reshape2::melt(matrix_pvalue), c("phenotype","module","nom_p"))
  matrix_rsquared_df = setNames(reshape2::melt(matrix_rsquared), c("phenotype","module","rsquared"))
  all_stats_df = matrix_pvalue_df %>% left_join(matrix_rsquared_df) %>% arrange(nom_p)

  return(list(all_stats_df = all_stats_df, matrix_rsquared = matrix_rsquared, matrix_pvalue = matrix_pvalue))
}

plot_module_trait_association_heatmap <- function(res_test, to_show, show_only_significant = F, signif_cutoff = c("***","**","*")){
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)

  matrix_rsquared = res_test$matrix_rsquared
  matrix_pvalue = res_test$matrix_pvalue
  
  matrix_rsquared_to_plot = matrix_rsquared[,to_show]
  matrix_pvalue_to_plot = matrix_pvalue[,to_show]
  
  # Adjust P-values by each phenotype separately.
  adj_matrix_pvalue_to_plot = matrix_pvalue_to_plot
  for(i in 1:nrow(matrix_pvalue_to_plot)){
    message("Adjusting P-values by Bonferroni correction by each phenotype separately. Double check if rows of res_test$matrix_pvalue are phenotypes and columns are your features (e.g. modules)")
    adj_matrix_pvalue_to_plot[i,] = p.adjust(matrix_pvalue_to_plot[i,], method = "bonferroni")
  }
  adj_matrix_pvalue_to_plot.signif <- symnum(adj_matrix_pvalue_to_plot, corr = FALSE, na = FALSE, 
                                             cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                                             symbols = c("***", "**", "*", ".", " "))
      
  log_matrix_pvalue_to_plot = -log10(matrix_pvalue_to_plot)
  dimnames(log_matrix_pvalue_to_plot) = dimnames(log_matrix_pvalue_to_plot)
  
  if(show_only_significant){
    if(is.numeric(signif_cutoff)){
      to_keep = colSums(adj_matrix_pvalue_to_plot <= signif_cutoff) > 0
    }else{
      to_keep = rep(F,ncol(adj_matrix_pvalue_to_plot.signif))
      for(cut_i in signif_cutoff){
        to_keep = to_keep | colSums(adj_matrix_pvalue_to_plot.signif == cut_i) > 0 # change for the significance you want 
      }
    }
    log_matrix_pvalue_to_plot = log_matrix_pvalue_to_plot[,to_keep]
    adj_matrix_pvalue_to_plot.signif = adj_matrix_pvalue_to_plot.signif[,to_keep]
  }
  
  matrix_pvalue_to_plot_labels = formatC(log_matrix_pvalue_to_plot, format = "f", digits = 2)
  log_matrix_pvalue_to_plot_t = t(log_matrix_pvalue_to_plot)
  
  # Colored by -log10(pvalue)
  # Numbers inside cell = -log10(pvalue): nominal
  Heatmap(log_matrix_pvalue_to_plot_t, name = "-log10(P-value)",
          cell_fun = function(j, i, x, y, width, height, fill) {
            if(as.character(t(adj_matrix_pvalue_to_plot.signif)[i,j]) == " "){
              grid.text( t(matrix_pvalue_to_plot_labels)[i,j], x, y, 
                        gp = gpar(fontsize = 8))
            }else{
              grid.text(paste0( t(matrix_pvalue_to_plot_labels)[i,j],"\n", t(adj_matrix_pvalue_to_plot.signif)[i,j] ), x, y, 
                        gp = gpar(fontsize = 8))
            }
          },
          col = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100),
          row_names_side = "left", show_row_names = T,
          cluster_rows = F, cluster_columns = F,
          column_names_gp = gpar(fontsize = 9),
          row_names_gp = gpar(fontsize = 9),
          show_row_dend = F, show_column_dend = F, rect_gp = gpar(col = "white", lwd = 1))
}
```

```{r Phenotypes_list}
pheno_list = c(# "cogn_global_lv"="gaussian", # Global cognitive function - Average of 19 tests
               "cogng_demog_slope"="gaussian", # Cognitive decline slope. Remove the effect of demog 
               "cogng_path_slope"="gaussian", # Resilience, removed the effect of path + demog
               "tangles_sqrt"="gaussian", # Tangle density - Mean of 8 brain regions
               # "nft_sqrt"="gaussian", # Neurofibrillary tangle summary based on 5 regions
               "amyloid_sqrt"="gaussian", # Overall amyloid level - Mean of 8 brain regions
               # "plaq_n_sqrt"="gaussian", # Neuritic plaque summary based on 5 regions
               # "plaq_d_sqrt"="gaussian", # Diffuse plaque summary based on 5 regions
               # "caa_4gp"="gaussian", # Cerebral amyloid angiopathy - 4 stages
               "gpath"="gaussian", # Global burden of AD pathology based on 5 regions
               "tdp_cs_6reg"="gaussian", # TDP-43, 6 region severity summary
               # "parksc_lv"="gaussian", # Global parkinsonian summary score
               # "cpd_lv"="binomial", # Clinical Parkinson's Disease 
               # "dxpark_status"="binomial", # Final Clinical Dx - Hx of Parkinson's disease/Parkinsonism (excl 9)
               "ad_dementia_status"="binomial", # Clinical AD # CT = MCI + NCI 
               "ci_status"="binomial" # AD+ MCI vs NCI 
               # "tdp_43_binary"="binomial",
               # "ci_num2_gct"="binomial", # Cerebral Infarctions - Binary - Gross-Chronic-Any Location
               # "ci_num2_mct"="binomial", # Cerebral Infarctions - Binary - Micro-Chronic-Any Location
               # "arteriol_scler"="gaussian", # Arteriolosclerosis - 4 stages
               # "cvda_4gp2"="gaussian", # Cerebral Atherosclerosis Rating - 4 levels (None - severe)
               # "vasc_3dis_sum_lv"="gaussian", # Cumulative vascular disease burden - Average of 3 items (ROS/MAP/MARS)
               # "vasc_risks_sum_lv"="gaussian", # Cumulative vascular disease risk factors - Average of 3 items
               # "CR_slope_lpm"="gaussian",
               # "CR_slope_lmm"="gaussian",
               # "CR_mean_level_lpm"="gaussian",
               # "CR_mean_level_lmm"="gaussian",
               # "age_death"="gaussian",
               # "msex"="binomial"
               )
```

```{r input_folders}
net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_net_MF/"
work_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/bulk_dlpfc/assoc_analysis/"
```

# Regressions
Same covariates tested for the ST and SN networks. Show all modules. Input: Average expression by module. 

Numbers and colors : -log10(nominal pvalue)

cutpoints: 

< 0.001 = *** 

0.01 = ** 

0.05 = *

0.1 = . 

1 = " " 


# All modules 
```{r layer_results1, message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
### Modules from SE
modules_file = read.table(paste0(net_dir, "geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

### eigengenes and average expression 
load(paste0(net_dir, "lv3_moduleEigengenes.Rdata"))
mod_eigengene = lv3_moduleEigengenes$eigengenes
mod_average = lv3_moduleEigengenes$averageExpr

mod_average$projid = gsub("(.*)_(.*)", "\\2", rownames(mod_eigengene)) #get the projid to match with phenotype data 
rownames(mod_average) = mod_average$projid
mod_average$projid = NULL

data4linear_reg <- mod_average # or mod_eigengene

load("/pastel/projects/spatial_t/pseudo_bulk/phenotypes.RData") # phenotypes
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid),]
all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs

res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ"))
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# save(res_test, file = paste0(work_dir, "results_lr_bulk_MF.Rdata"))
plot_module_trait_association_heatmap(res_test, to_show)
```

## Top results
Top result by covariate. 

```{r best_results9}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np}
createDT(matrix_pvalue)
```

# Significant results
## Adj P < 0.05
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r fig.height=10, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T, signif_cutoff = 0.05)
```

## Adj P < 0.000001

```{r fig.height=4, fig.width=5}
# pdf(paste0(net_dir, "bulk_DLPFC_topmod.pdf"), width = 5, height = 4)
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T, signif_cutoff = 0.000001)
# dev.off()
```

```{r}
res_test$all_stats_df$adj_pvalue = p.adjust(res_test$all_stats_df$nom_p, 
                                            method = "fdr")

createDT(res_test$all_stats_df)
```

# Session info
```{r}
sessionInfo()
```

