---
title: "Replication Analysis"
subtitle: "Eigengene calculation"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo A. Vialle
 |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Bulk RNASeq data from Mayo Clinic.  

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r Helper Functions, echo = FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}

## Run regression tests
run_module_trait_association <- function(data4linear_reg, # Matrix with module eigengenes (predictor)
                                         phenotype_dt, # Matrix with covariates (outcome + covariates)
                                         pheno_list, # List of phenotypes to be tested (with classes = binomial or gaussian)
                                         covariates = c("ageDeath","sex") # List of covariates to be adjusted
                                         ){ 
  
  if (!require("lme4")) install.packages("lme4")
  if (!require("lmerTest")) install.packages("lmerTest")
  if (!require("performance")) install.packages("performance")
  library(lme4)
  library(lmerTest)
  library(performance)
  
  outcome = names(pheno_list)
  outcome.family = pheno_list
  # random_effect = "projid"
  # avg_over_random_effect = T
  
  matrix_rsquared = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average)) #Number of modules
  matrix_pvalue = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  
  for (x in 1:length(pheno_list)){
    for (y in 1:ncol(mod_average)){
      outcome_pheno = outcome[x]
      outcome_type = outcome.family[x]

      dat4test_1 = setNames(as.data.frame(cbind(phenotype_dt[,outcome_pheno],data4linear_reg[,y])), c("outcome","predictor"))
      if(!is.null(covariates)){
        dat4test_2 = phenotype_dt[,covariates,drop=F]
        dat4test = na.omit(cbind(dat4test_1, dat4test_2))
        formula_string = as.formula(paste0("outcome ~ predictor + ", paste(covariates, collapse = " + ")))
        print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y], " + ", paste(covariates, collapse = " + ")))
      }else{
        dat4test = na.omit(dat4test_1)
        formula_string = as.formula(paste0("outcome ~ predictor"))
        print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y]))
      }
      
      if (outcome_type == "gaussian"){
        mod.obj0 = lm(formula_string, dat4test, na.action = "na.exclude")
        matrix_rsquared[x,y] <- summary( mod.obj0 )$adj.r.squared
        matrix_pvalue[x,y] <- summary( mod.obj0 )$coefficients["predictor","Pr(>|t|)"] #To insert pvalues in the heatmap
      }
      if (outcome_type == "binomial"){
        dat4test$outcome = as.factor(dat4test$outcome)
        mod.obj1 = glm(formula_string, dat4test, family = binomial, na.action = "na.exclude")
        matrix_rsquared[x,y] <-  1 - mod.obj1$deviance/mod.obj1$null.deviance # Pseudo r-squared
        matrix_pvalue[x,y] <- coef(summary(mod.obj1))["predictor",'Pr(>|z|)']
      }
    }
  }
  
  rownames(matrix_rsquared) = names(pheno_list)
  rownames(matrix_pvalue) = names(pheno_list)
  colnames(matrix_rsquared) = colnames(data4linear_reg)
  colnames(matrix_pvalue) = colnames(data4linear_reg)
  
  matrix_pvalue_df = setNames(reshape2::melt(matrix_pvalue), c("phenotype","module","nom_p"))
  matrix_rsquared_df = setNames(reshape2::melt(matrix_rsquared), c("phenotype","module","rsquared"))
  all_stats_df = matrix_pvalue_df %>% left_join(matrix_rsquared_df) %>% arrange(nom_p)

  return(list(all_stats_df = all_stats_df, matrix_rsquared = matrix_rsquared, matrix_pvalue = matrix_pvalue))
}

plot_module_trait_association_heatmap <- function(res_test, to_show, show_only_significant = F, signif_cutoff = c("***","**","*")){
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)

  matrix_rsquared = res_test$matrix_rsquared
  matrix_pvalue = res_test$matrix_pvalue
  
  matrix_rsquared_to_plot = matrix_rsquared[,to_show]
  matrix_pvalue_to_plot = matrix_pvalue[,to_show]
  
  # Adjust P-values by each phenotype separately.
  adj_matrix_pvalue_to_plot = matrix_pvalue_to_plot
  for(i in 1:nrow(matrix_pvalue_to_plot)){
    message("Adjusting P-values by Bonferroni correction by each phenotype separately. Double check if rows of res_test$matrix_pvalue are phenotypes and columns are your features (e.g. modules)")
    adj_matrix_pvalue_to_plot[i,] = p.adjust(matrix_pvalue_to_plot[i,], method = "bonferroni")
  }
  adj_matrix_pvalue_to_plot.signif <- symnum(adj_matrix_pvalue_to_plot, corr = FALSE, na = FALSE, 
                                             cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                                             symbols = c("***", "**", "*", ".", " "))
      
  log_matrix_pvalue_to_plot = -log10(matrix_pvalue_to_plot)
  dimnames(log_matrix_pvalue_to_plot) = dimnames(log_matrix_pvalue_to_plot)
  
  if(show_only_significant){
    if(is.numeric(signif_cutoff)){
      to_keep = colSums(adj_matrix_pvalue_to_plot <= signif_cutoff) > 0
    }else{
      to_keep = rep(F,ncol(adj_matrix_pvalue_to_plot.signif))
      for(cut_i in signif_cutoff){
        to_keep = to_keep | colSums(adj_matrix_pvalue_to_plot.signif == cut_i) > 0 # change for the significance you want 
      }
    }
    log_matrix_pvalue_to_plot = log_matrix_pvalue_to_plot[,to_keep]
    adj_matrix_pvalue_to_plot.signif = adj_matrix_pvalue_to_plot.signif[,to_keep]
  }
  
  matrix_pvalue_to_plot_labels = formatC(log_matrix_pvalue_to_plot, format = "f", digits = 2)
  log_matrix_pvalue_to_plot_t = t(log_matrix_pvalue_to_plot)
  
  # Colored by -log10(pvalue)
  # Numbers inside cell = -log10(pvalue): nominal
  Heatmap(log_matrix_pvalue_to_plot_t, name = "-log10(P-value)",
          cell_fun = function(j, i, x, y, width, height, fill) {
            if(as.character(t(adj_matrix_pvalue_to_plot.signif)[i,j]) == " "){
              grid.text( t(matrix_pvalue_to_plot_labels)[i,j], x, y, 
                        gp = gpar(fontsize = 8))
            }else{
              grid.text(paste0( t(matrix_pvalue_to_plot_labels)[i,j],"\n", t(adj_matrix_pvalue_to_plot.signif)[i,j] ), x, y, 
                        gp = gpar(fontsize = 8))
            }
          },
          col = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100),
          row_names_side = "left", show_row_names = T,
          cluster_rows = F, cluster_columns = F,
          column_names_gp = gpar(fontsize = 9),
          row_names_gp = gpar(fontsize = 9),
          show_row_dend = F, show_column_dend = F, rect_gp = gpar(col = "white", lwd = 1))
}
```

```{r Phenotypes_list}
# gaussian to run the linear regression 
pheno_list = c("Braak" = "gaussian",
               "thal" = "gaussian"
#               "diagnosis" = "binomial" # AD x CT 
)
```

```{r load_packages, message=F, warning=F}
work_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/bulk_dlpfc/replication/"
resources = "/pastel/resources/MayoClinic/RNAseq/"
net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_net_MF/"

library(tidyverse)
library(janitor)
library(WGCNA)
```

We already have the modules from our ROSMAP dataset, so I'll use this as gene list and calculate the eigengenes and module average expression for the same genes for Mayo. Then, we'll run regressions with their AD-related covariates to check for associations. 

# Input external dataset 

```{r data}
metadata = read.csv(paste0(resources, "RNAseq_Harmonization_Mayo_combined_metadata.csv"))
metadata$ageDeath = as.numeric(gsub("\\+", "", metadata$ageDeath))

message(paste0("Unique individualID: ", length(unique(metadata$individualID)))) 
message(paste0("Unique specimenID: ", length(unique(metadata$specimenID)))) 
table(metadata$tissue)
```

# Select samples
To avoid repeated measures. 
```{r sel}
# metadata 
metadata_temp = metadata[metadata$exclude == "FALSE", ] # Remove the samples that didn't pass QC 
# metadata_temp = metadata_temp[metadata_temp$diagnosis %in% c("Alzheimer Disease", "control"), ] # We filter to get only AD and CT
metadata_sel = metadata_temp[metadata_temp$tissue == "temporal cortex", ] # select the temporal cortex
# expression 
expr_mtx = read.table(paste0(resources, "Mayo_Normalized_counts_CQN.tsv"), check.names = F, header = T)
rownames(expr_mtx) = expr_mtx$feature
expr_mtx_sel = expr_mtx[, colnames(expr_mtx) %in% metadata_sel$specimenID] # select the same samples from the metadata 

```

# Input ROSMAP networks
Bulk DLPFC. 
```{r rosmap_modules}
modules_file = read.table(paste0(net_dir, "geneBycluster.txt"), header = T)

modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")
too_small = as.character( modules_size$module[modules_size$n_nodes < 30]) # Get modules < 30 nodes to be removed
net_output = modules_file %>% filter(! cluster_lv3 %in% as.integer(too_small)) # the clusters with at least 30 nodes 
net_output = net_output[,c("ensembl", "cluster_lv3", "gene_type", "symbol")]
net_output$ensembl2 = gsub("(.*)\\.(.*)", "\\1",net_output$ensembl)

message(paste0("Number of modules: ", length(unique(net_output$cluster_lv3))))
```

# Average expression - Mayo 
I got our gene list and calculated the ME for Mayo Clinic expression. 
```{r selec_genes}
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(net_output[,-1], by = c("ensembl" = "ensembl2"))

expr_mtx_sel_ord = expr_mtx_sel[gene_mod_map$ensembl,] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 
colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord_t = as.data.frame(t(expr_mtx_sel_ord))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# save results
# save(external_ME, metadata_sel, file = paste0(work_dir, "Mayo_moduleEigengenes_lv3.Rdata"))
# write.table(external_ME$eigengenes, file = paste0(work_dir, "Mayo_moduleEigengenes_lv3.txt"), sep = "\t", quote = F, row.names = T)
# write.table(external_ME$averageExpr, file = paste0(work_dir, "Mayo_moduleaverageExpr_lv3.txt"), sep = "\t", quote = F, row.names = T)

createDT(mod_average)
```

# Regressions 

Input: Average expression by module. 

Numbers and colors : -log10(nominal pvalue)

cutpoints: 

< 0.001 = *** 

0.01 = ** 

0.05 = *

0.1 = . 

1 = " " 

```{r data4reg}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = metadata_sel[match(rownames(data4linear_reg), metadata_sel$specimenID), ]
all(rownames(data4linear_reg) == phenotype_dt$specimenID) # Must be TRUE. Match IDs
# hist(phenotype_dt$Braak)

message(paste0("Unique individualID: ", length(unique(phenotype_dt$individualID)))) 
message(paste0("Unique specimenID: ", length(unique(phenotype_dt$specimenID)))) 
```

```{r reg1, message=FALSE, results='hide', warning=FALSE, fig.width=2.5, fig.height=10}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("ageDeath","sex")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf("/pastel/Github_scripts/SpeakEasy_dlpfc/figures4paper/reg_Mayo_adjcov.pdf", width = 2.5, height = 11)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```


## Top results
Top result by covariate. 

```{r best_results9}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np}
createDT(matrix_pvalue)
```

# Significant results
## Adj P < 0.05
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r fig.height=9, fig.width=2.5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T, signif_cutoff = 0.05)
```

## Adj P < 0.000001

```{r fig.height=5, fig.width=3}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T, signif_cutoff = 0.000001)
```

```{r}
res_test$all_stats_df$adj_pvalue = p.adjust(res_test$all_stats_df$nom_p, 
                                            method = "fdr")

createDT(res_test$all_stats_df)
```

# Session info
```{r}
sessionInfo()
```
