---
title: "Single nuclei from DLPFC"
subtitle: "Regressions between the modules and the covariates"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo Vialle 
 | 
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
params:
  cell_type: "mic"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

> Clusters from level 03 of SE | The regressions are adjusted for age, sex and educ. 

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(ggeasy)
library(ggplot2)
library(broom)
library(gplots)
library(ggpubr)
library(tidyverse)
```

```{r Helper Functions}
## Create table 
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}

## Run regression tests
run_module_trait_association <- function(data4linear_reg, # Matrix with module eigengenes (predictor)
                                         phenotype_dt, # Matrix with covariates (outcome + covariates)
                                         pheno_list, # List of phenotypes to be tested (with classes = binomial or gaussian)
                                         covariates = c("age_death","msex", "educ"), # List of covariates to be adjusted
                                         verbose = T){ 
  
  if (!require("lme4")) install.packages("lme4")
  if (!require("lmerTest")) install.packages("lmerTest")
  if (!require("performance")) install.packages("performance")
  library(lme4)
  library(lmerTest)
  library(performance)
  
  outcome = names(pheno_list)
  outcome.family = pheno_list

  matrix_rsquared = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average)) #Number of modules
  matrix_pvalue = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  matrix_tstats = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  
  for (x in 1:length(pheno_list)){
    for (y in 1:ncol(mod_average)){
      # x = 1
      # y = 1
      outcome_pheno = outcome[x]
      outcome_type = outcome.family[x]

      dat4test_1 = setNames(as.data.frame(cbind(phenotype_dt[,outcome_pheno],data4linear_reg[,y])), c("outcome","predictor"))
      if(!is.null(covariates)){
        dat4test_2 = phenotype_dt[,covariates,drop=F]
        dat4test = na.omit(cbind(dat4test_1, dat4test_2))
        formula_string = as.formula(paste0("outcome ~ predictor + ", paste(covariates, collapse = " + ")))
        if(verbose) print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y], " + ", paste(covariates, collapse = " + ")))
      }else{
        dat4test = na.omit(dat4test_1)
        formula_string = as.formula(paste0("outcome ~ predictor"))
        if(verbose) print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y]))
      }
      
      if (outcome_type == "gaussian"){
        mod.obj0 = lm(formula_string, dat4test, na.action = "na.exclude")
        matrix_rsquared[x,y] <- summary( mod.obj0 )$adj.r.squared
        matrix_pvalue[x,y] <- summary( mod.obj0 )$coefficients["predictor","Pr(>|t|)"] #To insert pvalues in the heatmap
        matrix_tstats[x,y] <- summary( mod.obj0 )$coefficients["predictor","t value"]
      }
      if (outcome_type == "binomial"){
        dat4test$outcome = as.factor(dat4test$outcome)
        mod.obj1 = glm(formula_string, dat4test, family = binomial, na.action = "na.exclude")
        matrix_rsquared[x,y] <-  1 - mod.obj1$deviance/mod.obj1$null.deviance # Pseudo r-squared
        matrix_pvalue[x,y] <- coef(summary(mod.obj1))["predictor",'Pr(>|z|)']
        matrix_tstats[x,y] <- summary( mod.obj1 )$coefficients["predictor","z value"]
      }
    }
  }
  
  rownames(matrix_rsquared) = names(pheno_list)
  rownames(matrix_pvalue) = names(pheno_list)
  colnames(matrix_rsquared) = colnames(data4linear_reg)
  colnames(matrix_pvalue) = colnames(data4linear_reg)
  rownames(matrix_tstats) = names(pheno_list)
  colnames(matrix_tstats) = colnames(data4linear_reg)
  
  matrix_pvalue_df = setNames(reshape2::melt(matrix_pvalue), c("phenotype","module","nom_p"))
  matrix_rsquared_df = setNames(reshape2::melt(matrix_rsquared), c("phenotype","module","rsquared"))
  matrix_tstats_df = setNames(reshape2::melt(matrix_tstats), c("phenotype","module","tstats"))
  all_stats_df = matrix_pvalue_df %>% left_join(matrix_rsquared_df) %>% left_join(matrix_tstats_df) %>%  arrange(nom_p)

  return(list(all_stats_df = all_stats_df, matrix_rsquared = matrix_rsquared, matrix_pvalue = matrix_pvalue, matrix_tstats = matrix_tstats_df))
}


plot_module_trait_association_heatmap <- function(res_test, to_show, show_only_significant = F, signif_cutoff = c("***","**","*")){
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)

  matrix_rsquared = res_test$matrix_rsquared
  matrix_pvalue = res_test$matrix_pvalue
  
  matrix_rsquared_to_plot = matrix_rsquared[,to_show]
  matrix_pvalue_to_plot = matrix_pvalue[,to_show]
  
  # Adjust P-values by each phenotype separately.
  adj_matrix_pvalue_to_plot = matrix_pvalue_to_plot
  for(i in 1:nrow(matrix_pvalue_to_plot)){
    adj_matrix_pvalue_to_plot[i,] = p.adjust(matrix_pvalue_to_plot[i,], method = "bonferroni")
  }
  adj_matrix_pvalue_to_plot.signif <- symnum(adj_matrix_pvalue_to_plot, corr = FALSE, na = FALSE, 
                                             cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                                             symbols = c("***", "**", "*", ".", " "))
      
  log_matrix_pvalue_to_plot = -log10(matrix_pvalue_to_plot)
  dimnames(log_matrix_pvalue_to_plot) = dimnames(log_matrix_pvalue_to_plot)
  
  if(show_only_significant){
    if(is.numeric(signif_cutoff)){
      to_keep = colSums(adj_matrix_pvalue_to_plot <= signif_cutoff) > 0
    }else{
      to_keep = rep(F,ncol(adj_matrix_pvalue_to_plot.signif))
      for(cut_i in signif_cutoff){
        to_keep = to_keep | colSums(adj_matrix_pvalue_to_plot.signif == cut_i) > 0 # change for the significance you want 
      }
    }
    log_matrix_pvalue_to_plot = log_matrix_pvalue_to_plot[,to_keep,drop=F]
    adj_matrix_pvalue_to_plot.signif = adj_matrix_pvalue_to_plot.signif[,to_keep,drop=F]
  }
  
  matrix_pvalue_to_plot_labels = formatC(log_matrix_pvalue_to_plot, format = "f", digits = 2)
  log_matrix_pvalue_to_plot_t = t(log_matrix_pvalue_to_plot)
  
  # Colored by -log10(pvalue)
  # Numbers inside cell = -log10(pvalue): nominal
  Heatmap(log_matrix_pvalue_to_plot_t, name = "-log10(P-value)",
          cell_fun = function(j, i, x, y, width, height, fill) {
            if(as.character(t(adj_matrix_pvalue_to_plot.signif)[i,j]) == " "){
              grid.text( t(matrix_pvalue_to_plot_labels)[i,j], x, y, 
                        gp = gpar(fontsize = 8))
            }else{
              grid.text(paste0( t(matrix_pvalue_to_plot_labels)[i,j],"\n", t(adj_matrix_pvalue_to_plot.signif)[i,j] ), x, y, 
                        gp = gpar(fontsize = 8))
            }
          },
          col = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100),
          row_names_side = "left", show_row_names = T,
          cluster_rows = F, cluster_columns = F,
          column_names_gp = gpar(fontsize = 9),
          row_names_gp = gpar(fontsize = 9),
          show_row_dend = F, show_column_dend = F, rect_gp = gpar(col = "white", lwd = 1))
}
```

```{r Phenotypes_list}
pheno_list = c("cogng_demog_slope"="gaussian", # Cognitive decline slope. Remove the effect of demog 
               "cogng_path_slope"="gaussian", # Resilience, removed the effect of path + demog
               "tangles_sqrt"="gaussian", # Tangle density - Mean of 8 brain regions
               "amyloid_sqrt"="gaussian", # Overall amyloid level - Mean of 8 brain regions
               "gpath"="gaussian", # Global burden of AD pathology based on 5 regions
               "tdp_cs_6reg"="gaussian", # TDP-43, 6 region severity summary
               "ad_dementia_status"="binomial" # Clinical AD # CT = MCI + NCI 
               )
```


```{r input_folders}
net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/results/"
non_modules_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/figures4paper/v2_mar2024/"
work_dir = tempdir()
```

# Run regressions
Input: Average expression by module. 

```{r regressions, message=FALSE, results='hide', warning=FALSE}
load("/pastel/projects/spatial_t/pseudo_bulk/phenotypes.RData") # phenotypes

all_stats = data.frame()
for(cell_i in c("ext","inh","ast","oli","mic","end","opc")){
  modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
  modules_size = as.data.frame( table(modules_file$cluster_lv3))
  colnames(modules_size) = c("module", "n_nodes")
  
  ### eigengenes and average expression 
  load(paste0(net_dir, cell_i,"/lv3_moduleEigengenes.Rdata"))
  mod_eigengene = lv3_moduleEigengenes$eigengenes
  mod_average = lv3_moduleEigengenes$averageExpr
  
  mod_average$projid = gsub("(.*)_(.*)", "\\2", rownames(mod_eigengene)) #get the projid to match with phenotype data 
  rownames(mod_average) = mod_average$projid
  mod_average$projid = NULL
  
  data4linear_reg <- mod_average # or mod_eigengene
  
  phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid),]
  all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
  
  res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ"), verbose = F)
  matrix_rsquared = res_test$matrix_rsquared
  matrix_pvalue = res_test$matrix_pvalue
  save(res_test, file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))
  
  stats_df = res_test$all_stats_df
  stats_df$network = cell_i
  all_stats = rbind(all_stats, stats_df)
}

save(all_stats, file = paste0(work_dir, "all_res_test_stats_SN.Rdata"))
```


# Complete table
Remove the non-modules. 
```{r table}
load(paste0(work_dir, "all_res_test_stats_SN.Rdata"))

# Non-modules to be removed (< 30 nodes)
emods2remove = read.table(paste0(non_modules_dir, "non_modules.txt"), header = T, stringsAsFactors = F)

all_stats$module_temp <- gsub("AE", "M", all_stats$module)
all_stats$module2 <- paste0(all_stats$network, "_", all_stats$module_temp)
all_stats$module_temp <- NULL

# Remove non-modules
all_stats_filt <- all_stats[! all_stats$module2 %in% emods2remove$module2, ] 
# dim(all_stats_filt) # 1351    6
# length(unique(all_stats_filt$module2)) # 193
# length(unique(all_stats_filt$phenotype)) # 7

# Adjust by FDR
all_stats_filt$FDR = p.adjust(all_stats_filt$nom_p, method = "fdr")
write_tsv(all_stats_filt, "/pastel/Github_scripts/SpeakEasy_dlpfc/sn_dlpfc/2nd_pass/snModules_ADtraits_results.tsv.gz")

createDT(all_stats_filt %>% arrange(nom_p))
```

# Best hit per module
```{r}
# select best hit of each module
best_hit_mod_assoc = all_stats_filt %>% filter(FDR < 0.05) %>% 
  group_by(module2) %>% slice_min(order_by = FDR, n = 1) 

createDT(best_hit_mod_assoc)
```

# Number of significant modules per network

Adjusted by ALL modules and phenotypes. FDR < 0.05

```{r}
# count number of signif modules per network (at least one trait)
best_hit_mod_assoc %>% dplyr::select(network,module2) %>% distinct() %>% group_by(network) %>% dplyr::summarise(n = n()) %>%
  ggplot(aes(x = network, y = n)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = n), vjust = 1.5, color = "white") +
  theme_classic() +
  labs(title = "Number of significant modules per network")
```

# Exploratory visualizations by cell type

Heatmaps

Numbers and colors : -log10(nominal pvalue)

Cutpoints (adjusted pvalue by bonferroni, by phenotype)

< 0.001 = *** 

0.01 = ** 

0.05 = *

0.1 = . 

1 = " " 

# Ext 
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "ext"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results1}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np1}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig1, fig.height=8, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# Inh
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "inh"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results2}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np2}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig2, fig.height=6, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# Oli
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "oli"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results3}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np3}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig3, fig.height=3.5, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# End
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "end"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results4}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np4}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig4, fig.height=3, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# Ast
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "ast"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results5}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np5}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig5, fig.height=4, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# Mic
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "mic"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results6}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np6}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig6, fig.height=3, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```

# OPCs
## All modules 
```{r message=FALSE, results='hide', warning=FALSE, fig.width=5, fig.height=11}
cell_i = "opc"
### Modules from SE
modules_file = read.table(paste0(net_dir, cell_i, "/geneBycluster.txt"), header = T) 
modules_size = as.data.frame( table(modules_file$cluster_lv3))
colnames(modules_size) = c("module", "n_nodes")

load(file = paste0(work_dir, "results_lr_",cell_i,".Rdata"))

### Get modules >= 30 nodes to show:
to_show = paste0("AE",as.character( modules_size$module[modules_size$n_nodes >= 30])) 

# pdf(paste0(work_dir, "assoc_ext_adj.pdf"), width = 5, height = 10)
plot_module_trait_association_heatmap(res_test, to_show)
# dev.off()
```

## Top results
Top result by covariate. 

```{r best_results7}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

## Nominal pvalue
```{r np7}
createDT(matrix_pvalue)
```

## Significant results 
Threshold: At least one module with adjusted pvalue < 0.05. 

Method: Bonferroni by column. 
```{r sig7, fig.height=3.5, fig.width=5}
plot_module_trait_association_heatmap(res_test, to_show, show_only_significant = T)
```


# Session info
```{r}
sessionInfo()
```
