---
title: "Single nuclei RNASeq from DLPFC"
subtitle: "Phenotype data"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo A. Vialle
 | Shinya Tasaki
 |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(ggplot2)
library(ggeasy)
library(ggpubr)
library(kableExtra)
library(gridExtra)
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

```{r}
require(tidyverse)

# cleanProjidsInDataTable
#
# = input =
# datatable: two-dimensional table, such as imported by tidyverse::read_csv
# projid_columns: vector of column names, each of which should appear exactly once in datatable
#
# = output =
# a copy of the datatable where the projid_columns have each been cleaned
# into a vector of character strings exactly eight digits
cleanProjidsInDataTable = function(datatable, projid_columns) {
  copytable = data.table::copy(datatable)
  for (column in projid_columns) {
    selection = select(datatable, all_of(column))
    if (length(selection) == 1) {
      copytable[[column]] = cleanProjidsInVector(selection[[column]])
    } else {
      stop(paste("Unique '", column, "' column not found in datatable.", sep="", collapse=NULL))
    }  
  }
  return(copytable)
}

cleanProjidsInVector = function(values) {
  if (is.integer(values)) {
    return(sapply(values, cleanIntegerProjid))
  } else if (is.double(values)) {
    return(sapply(values, cleanDoubleProjid))
  } else if (is.character(values)) {
    return(sapply(values, cleanStringProjid))
  } else {
    stop("Invalid vector type for cleaning projids.")
  }
}

cleanIntegerProjid = function(value) {
  if (is.integer(value)) {
    if (is.na(value)) {
      return(NA)
    } else if (value >= 0 && value <= 99999999) {
      string = as.character(value)
      zeroes = createLeadingZeroes(as.integer(8)-nchar(string))
      return(paste(zeroes, string, sep="", collapse=""))
    } else {
      stopCleaningProjid(value)
    }  
  } else {
    stop("Attempted to clean non-integer value as integer projid.")
  }
}

createLeadingZeroes = function(count) {
  if (is.integer(count) && count >= 0) {
    return(paste(rep("0", count), collapse=""))
  } else {
    stop(paste("Invalid count of leading zeroes: ", count, sep="", collapse=NULL))
  }
}

cleanDoubleProjid = function(value) {
  if (is.double(value)) {
    if (is.na(value)) {
      return(NA)
    } else if (value == as.integer(value)) {
      return(cleanIntegerProjid(as.integer(value)))
    } else {
      stopCleaningProjid(value)
    }  
  } else {
    stop("Attempted to clean non-double value as double projid.")
  }
}

cleanStringProjid = function(value) {
  if (is.character(value)) {
    if (is.na(value)) {
      return(NA)
    } else if (str_detect(value, "^\\d{1,8}$")) {
      return(cleanIntegerProjid(as.integer(value)))
    } else {
      stopCleaningProjid(value)
    }
  } else {
    stop("Attempted to clean non-character value as character projid.")
  }
}

stopCleaningProjid = function(projid) {
  stop(paste("Invalid projid value for cleaning: ", value, sep="", collapse=NULL))
}
```

```{r input}
resource_dir = "/pastel/resources/20220203_snRNAseq_AMPAD/"
annotation = readRDS(paste0(resource_dir,"updated_annotations/annotations.rds")) # 1638882 rows
annotation = cleanProjidsInDataTable(annotation,"projid")

load(paste0(resource_dir, "updated_annotations/celltype_exp.RData"))
annotation.filt = annotation[annotation$projid %in% pheno.filt$projid,]
# pheno.filt
```

# Number of donors: 
```{r n_d}
length(unique(pheno.filt$projid))
length(unique(annotation.filt$projid))
```

# Final diagnosis 
Final consensus cognitive diagnosis: **cogdx**. Clinical consensus diagnosis of cognitive status at time of death. 
```{r cogdx}
counts_cogdx = as.data.frame(table(pheno.filt$cogdx))
colnames(counts_cogdx) = c("cogdx", "frequency")
counts_cogdx$cogdx_description = c("1 NCI: No cognitive impairment",
                                   "2 MCI: Mild cognitive impairment (One impaired domain) and NO other cause of CI",
                                   "3 MCI: Mild cognitive impairment (One impaired domain) AND another cause of CI",
                                   "4 AD: Alzheimer’s dementia and NO other cause of CI",
                                   "5 AD: Alzheimer’s dementia AND another cause of CI",
                                   "6 Other dementia: Other primary cause of dementia")

createDT(counts_cogdx)
```

# Cell types: 
## Class 
```{r counts_cells}
createDT(as.tibble(unique(annotation.filt[, c("cell.type", "class")])))
```

## Cells by donor  
What is the total number of cells by donor? 
```{r cell_counts_d, message=FALSE, warning=FALSE}
length(unique(annotation.filt$projid))
cell_perc <- annotation.filt %>% group_by(projid, cell.type) %>% summarise(n_cells = n()) %>% ungroup() %>% group_by(projid) %>% mutate(perc_cells = 100*(n_cells/sum(n_cells)))

cell_perc_wide <- cell_perc %>% select(projid,cell.type,perc_cells) %>% pivot_wider(values_from = perc_cells, names_from = cell.type) # cell proportions by donor 
#write.table(cell_perc_wide, file = paste0(resource_dir, "updated_annotations/cellprop_snRNAseq.txt"), quote = F, row.names = F, sep = "\t")

subcell_perc <- annotation.filt %>% group_by(projid, state) %>% summarise(n_cells = n()) %>% ungroup() %>% group_by(projid) %>% mutate(perc_cells = 100*(n_cells/sum(n_cells)))
subcell_perc_wide <- subcell_perc %>% select(projid,state,perc_cells) %>% pivot_wider(values_from = perc_cells, names_from = state) # subcell proportions by donor 
#write.table(subcell_perc_wide, file = paste0(resource_dir, "updated_annotations/subcellprop_snRNAseq.txt"), quote = F, row.names = F, sep = "\t")

cellsBydonor = annotation.filt %>% group_by(projid) %>% summarise(n_cells = n()) %>% arrange(n_cells)
createDT(cellsBydonor)
```

```{r cell_counts_plot, fig.width=10, fig.height=3.5, dpi=300, message=FALSE}
cell_perc$projid = as.character(cell_perc$projid)
cell_perc_diag = cell_perc %>% left_join(pheno.filt[, c("projid", "cogdx")] )
cell_perc_diag$cogdx_group = as.character(cell_perc_diag$cogdx)
cell_perc_diag$cogdx_group[cell_perc_diag$cogdx_group %in% c("1")] <- "NCI" # Normal cognition
cell_perc_diag$cogdx_group[cell_perc_diag$cogdx_group %in% c("2", "3")] <- "MCI"
cell_perc_diag$cogdx_group[cell_perc_diag$cogdx_group %in% c("4", "5", "6")] <- "AD"

cell_color_map = list(`Excitatory Neurons` = ggsci::pal_d3("category10")(10)[1],
                      `Inhibitory Neurons` = ggsci::pal_d3("category10")(10)[2],
                      Oligodendrocytes = ggsci::pal_d3("category10")(10)[3],
                      Astrocyte = ggsci::pal_d3("category10")(10)[4],
                      Microglia = ggsci::pal_d3("category10")(10)[5],
                      OPCs = ggsci::pal_d3("category10")(10)[6],
                      Endothelial = ggsci::pal_d3("category10")(10)[9],
                      Other = "gray")

cell_labels = list(ext = "Excitatory Neurons",
                   inh = "Inhibitory Neurons",
                   oli = "Oligodendrocytes",
                   ast = "Astrocyte",
                   mic = "Microglia",
                   opc = "OPCs",
                   end = "Endothelial",
                   other = "Other")

cell_perc_diag$cell.type_group = cell_perc_diag$cell.type
cell_perc_diag$cell.type_group[! cell_perc_diag$cell.type_group %in% as.character(names(cell_color_map))] <- "Other"

donor_order <- cell_perc_diag %>% filter(cell.type=="Excitatory Neurons") %>% select(projid, perc_cells) 
donor_order <- donor_order[order(donor_order$perc_cells, decreasing = T),]
donor_order$n_order = 1:nrow(donor_order)

cell_perc_diag <- cell_perc_diag %>% left_join(donor_order[,c("projid","n_order")])

ggplot(cell_perc_diag, aes(x = reorder(projid, n_order, sum), y=perc_cells, fill=cell.type_group)) +
  geom_bar(stat = "identity") +
  facet_wrap(~cogdx_group, scales = "free_x") +
  scale_fill_manual(values=cell_color_map) +
  theme_classic() + 
  easy_remove_x_axis() + 
  labs(x = "Donor", y = "% of cells", fill = "Cell type") 
```

## Cells proportion 
```{r cell_prop}
cell_prop_total = annotation.filt %>% group_by(cell.type) %>% summarise(n_cells = n()) %>% mutate(perc_cells = n_cells/sum(n_cells)) %>%
  arrange(-perc_cells) %>% select(cell.type, perc_cells) %>% mutate(perc_cells = scales::percent(perc_cells))

createDT(cell_prop_total)
```

## Class proportion 
```{r class_prop}
class_prop_total = annotation.filt %>% group_by(class) %>% summarise(n_cells = n()) %>% mutate(perc_cells = n_cells/sum(n_cells)) %>%
  arrange(-perc_cells) %>% select(class, perc_cells) %>% mutate(perc_cells = scales::percent(perc_cells))

createDT(class_prop_total)
```

# Demographics
## Sex
```{r sex}
counts_sex = as.data.frame( table(pheno.filt$msex))
colnames(counts_sex) = c("Sex", "Frequency")
counts_sex$Sex = as.character(counts_sex$Sex)
counts_sex$Sex[counts_sex$Sex == 1] = "Male"
counts_sex$Sex[counts_sex$Sex == 0] = "Female"

kable(counts_sex) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"))
```

## Age range 
Variable: age at death.
```{r age_range, message=TRUE, fig.width=6, fig.height=4}
message(paste0("Age max: ", max(pheno.filt$age_death)))
message(paste0("Age min: ", min(pheno.filt$age_death)))

ageByDonor = unique(pheno.filt[,c("projid", "age_death", "msex")])
ageByDonor$msex = as.character(ageByDonor$msex)
ageByDonor$msex[ageByDonor$msex == 1] = "Male"
ageByDonor$msex[ageByDonor$msex == 0] = "Female"

mean_f = mean(ageByDonor[ageByDonor$msex == "Female", "age_death"], na.rm = T)
mean_m = mean(ageByDonor[ageByDonor$msex == "Male", "age_death"], na.rm = T)

message(paste0("Age mean female: ", mean_f))
message(paste0("Age mean male: ", mean_m))

ggplot(ageByDonor, aes(x=age_death, fill=msex)) +
  geom_histogram(bins = 25, colour='black', position = "stack") +
  labs(x="Age", y="Donors") +
  # scale_y_continuous(breaks = (1:20)) +
  # scale_x_continuous(breaks=seq(20,120,10)) + 
  geom_vline(xintercept=mean_f, color = "red", linetype="dashed") +
  geom_vline(xintercept=mean_m, color = "blue", linetype="dashed") +
  theme_classic()
```

# Plots by diagnosis
cogdx = Clinical consensus diagnosis of cognitive status at time of death. 

## Cognitive decline 
Variable: **cogng_random_slope**. Estimated slope from random effects model for global cognition. 
```{r slope, fig.width=10, fig.height=6, warning=FALSE}
# head(pheno.filt[,c("projid", "amyloid", "cogdx", "tangles", "cogng_random_slope")])
ggplot(data = pheno.filt, aes(x=projid, y = cogng_random_slope)) + 
  geom_point() +
  facet_wrap(~cogdx) +
  easy_remove_x_axis() +
  labs(x = "Individual", y = "Slope for global cognition")
```

## Amyloid
Overall **amyloid** level - Mean of 8 brain regions. 
```{r amyloid, fig.width=10, fig.height=6, warning=FALSE}
ggplot(data = pheno.filt, aes(x=projid, y = amyloid)) + 
  geom_point() +
  facet_wrap(~cogdx) +
  easy_remove_x_axis() +
  labs(x = "Individual", y = "Amyloid level")
```

## Tangles
**Tangle** density - Mean of 8 brain regions.
```{r tangles, fig.width=10, fig.height=6, warning=FALSE}
ggplot(data = pheno.filt, aes(x=projid, y = tangles)) + 
  geom_point() +
  facet_wrap(~cogdx) +
  scale_y_continuous(breaks = seq(0, 60, by=10)) +
  easy_remove_x_axis() +
  labs(x = "Individual", y = "Tangles density") 
```

# Plots for cogdx_group
T test for pairwise comparisons. 

Anova for global p-value.
```{r cogdx_group, fig.width=12, fig.height=6, warning=FALSE, dpi=300}
pheno.filt$cogdx_3gp_desc = pheno.filt$cogdx_3gp
pheno.filt$cogdx_3gp_desc[pheno.filt$cogdx_3gp_desc == 1] <- "NCI"
pheno.filt$cogdx_3gp_desc[pheno.filt$cogdx_3gp_desc == 2] <- "MCI"
pheno.filt$cogdx_3gp_desc[pheno.filt$cogdx_3gp_desc == 3] <- "AD"
  
my_comparisons = list(c("NCI", "AD"), c("NCI", "MCI"), c("MCI", "AD"))

p1 <- ggplot(pheno.filt, aes(x = cogdx_3gp_desc, y = cogng_random_slope)) + 
  geom_boxplot(notch = F, na.rm = T) + 
  stat_compare_means(comparisons = my_comparisons, method = "t.test")+
  stat_compare_means(method = "anova", label.y = 0.4) + 
  geom_jitter() +
  labs(x = "Diagnosis of cognitive status", y = "Slope for global cognition") +
  theme_classic()
  
p2 <- ggplot(pheno.filt, aes(x = cogdx_3gp_desc, y = amyloid)) + 
  geom_boxplot(notch = F, na.rm = T) + 
  stat_compare_means(comparisons = my_comparisons, method = "t.test")+
  stat_compare_means(method = "anova", label.y = 30) + 
  geom_jitter() +
  labs(x = "Diagnosis of cognitive status", y = "Amyloid levels") +
  theme_classic()

p3 <- ggplot(pheno.filt, aes(x = cogdx_3gp_desc, y = tangles)) + 
  geom_boxplot(notch = F, na.rm = T) + 
  stat_compare_means(comparisons = my_comparisons, method = "t.test")+
  stat_compare_means(method = "anova", label.y = 90) + 
  geom_jitter() +
  labs(x = "Diagnosis of cognitive status", y = "Tangles density") +
  theme_classic()

grid.arrange(p1, p2, p3, ncol=3)
```

# Session Info
```{r}
sessionInfo()
```

