---
title: "Single nuclei from DLPFC"
subtitle: "Module Eigengene and Average expression"
author:  |
 | Katia de Paiva Lopes 
 |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
params:
  cell_type: "mic"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
# rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(tidyverse)
library(janitor)
library(readxl)
library(WGCNA)

net_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/results/"
work_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/"
expression_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/input/"
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}
```

## Input 
In order to summarize the SE network information we will perform the module eigengene approach. Module eigengene is defined as the first principal component of the expression matrix of the corresponding module.

```{r input}
macro_type = params$cell_type #macro_structure. It can be cell_type, metabolites, region of the brain. 
message(paste0("Cell type: ", macro_type))

# Expression data for a single set in the form of a data frame where rows are samples and columns are genes (probes):
exprData = read.table(paste0(expression_dir, macro_type, ".txt"), header = T, stringsAsFactors = F, check.names = F) # Results of the expression
expr_matx_t = as.data.frame( t(exprData)) # Residuals of the expression

# Gene by module
gene_mod = read.table(paste0(net_dir, macro_type, "/geneBycluster.txt"), header = T, stringsAsFactors = F) # clusters from SpeakEasy

k_dataset_ord = gene_mod[match(colnames(expr_matx_t), gene_mod$ensembl), ] #order the matrices 
all(colnames(expr_matx_t) == k_dataset_ord$ensembl) # must be true 

```

## Module eigengenes - Lv3
Showing the module eigengenes in a dataframe, with each column corresponding to one eigengene. 
```{r eigen}
colors = k_dataset_ord$cluster_lv3
lv3_moduleEigengenes = moduleEigengenes(expr_matx_t, colors, verbose = 0)

# save results
save(lv3_moduleEigengenes, file = paste0(net_dir, macro_type,"/", "lv3_moduleEigengenes.Rdata"))
write.table(lv3_moduleEigengenes$eigengenes, file = paste0(net_dir, macro_type,"/", "lv3_moduleEigengenes.txt"), sep = "\t", quote = F, row.names = T)

createDT(lv3_moduleEigengenes$eigengenes)
```

## Average expression - Lv3
A dataframe containing average normalized expression in each module. 
```{r ave}
# The columns are named by the corresponding color with an "AE" prepended, e.g., AEturquoise etc.
colnames(lv3_moduleEigengenes$averageExpr) = gsub("AE", "AE_M", colnames(lv3_moduleEigengenes$averageExpr))
write.table(lv3_moduleEigengenes$averageExpr, file = paste0(net_dir, macro_type,"/", "lv3_averageExpr.txt"), sep = "\t", quote = F, row.names = T)
createDT(lv3_moduleEigengenes$averageExpr)
```

---

## Session info

```{r}
sessionInfo()
```
