---
title: "Single nuclei from DLPFC"
subtitle: "Replication vs MIT data"
author:  |
 | Katia de Paiva Lopes 
 | Ricardo A. Vialle
 | Shinya Tasaki
 |
 | Rush Alzheimer’s Disease Center
 | Chicago, IL 60612
date: "`r Sys.Date()`"
output: 
 rmarkdown::html_document:
   theme: spacelab
   highlight: tango
   code_folding: hide
   toc: true
   toc_float: true
   df_print: paged
   smooth_scroll: true
   number_sections: false
   self_contained: true  
---

```{r clean.variables, echo=FALSE}
#This command clean all variables. BE CAREFULL!!! 
rm(list = setdiff(ls(), lsf.str()))
```

```{r load.packages, echo=FALSE, message=FALSE, results='hide', warning=FALSE}
library(Seurat)
library(tidyverse)
library(kableExtra)
library(furrr)
library(future)
library(janitor)
library(WGCNA)
#future::plan(multiprocess)

knitr::opts_chunk$set( 
  warning=FALSE,
  message=FALSE)
```

```{css zoom-lib-src, echo = FALSE, eval=F}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

```{r Helper Functions, echo=FALSE}
createDT <- function(DF, caption="", scrollY=500){
  data <- DT::datatable(DF, caption=caption,
    extensions =  'Buttons',
    options = list( dom = 'Bfrtip', 
                    buttons = c('copy', 'csv', 'excel', 'pdf', 'print'), 
                    scrollY = scrollY, scrollX=T, scrollCollapse = T, paging = F,  
                      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    )
  ) 
   return(data)
}

## Run regression tests
run_module_trait_association <- function(data4linear_reg, # Matrix with module eigengenes (predictor)
                                         phenotype_dt, # Matrix with covariates (outcome + covariates)
                                         pheno_list, # List of phenotypes to be tested (with classes = binomial or gaussian)
                                         covariates = c("age_death","msex", "educ"), # List of covariates to be adjusted
                                         verbose = T){ 
  
  if (!require("lme4")) install.packages("lme4")
  if (!require("lmerTest")) install.packages("lmerTest")
  if (!require("performance")) install.packages("performance")
  library(lme4)
  library(lmerTest)
  library(performance)
  
  outcome = names(pheno_list)
  outcome.family = pheno_list

  matrix_rsquared = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average)) #Number of modules
  matrix_pvalue = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  matrix_tstats = matrix(NA, nrow = length(outcome), ncol = ncol(mod_average))
  
  for (x in 1:length(pheno_list)){
    for (y in 1:ncol(mod_average)){
      # x = 1
      # y = 1
      outcome_pheno = outcome[x]
      outcome_type = outcome.family[x]

      dat4test_1 = setNames(as.data.frame(cbind(phenotype_dt[,outcome_pheno],data4linear_reg[,y])), c("outcome","predictor"))
      if(!is.null(covariates)){
        dat4test_2 = phenotype_dt[,covariates,drop=F]
        dat4test = na.omit(cbind(dat4test_1, dat4test_2))
        formula_string = as.formula(paste0("outcome ~ predictor + ", paste(covariates, collapse = " + ")))
        if(verbose) print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y], " + ", paste(covariates, collapse = " + ")))
      }else{
        dat4test = na.omit(dat4test_1)
        formula_string = as.formula(paste0("outcome ~ predictor"))
        if(verbose) print(paste0("Testing (n=",nrow(dat4test),"):  ", outcome_pheno, " ~ ", names(data4linear_reg)[y]))
      }
      
      if (outcome_type == "gaussian"){
        mod.obj0 = lm(formula_string, dat4test, na.action = "na.exclude")
        matrix_rsquared[x,y] <- summary( mod.obj0 )$adj.r.squared
        matrix_pvalue[x,y] <- summary( mod.obj0 )$coefficients["predictor","Pr(>|t|)"] #To insert pvalues in the heatmap
        matrix_tstats[x,y] <- summary( mod.obj0 )$coefficients["predictor","t value"]
      }
      if (outcome_type == "binomial"){
        dat4test$outcome = as.factor(dat4test$outcome)
        mod.obj1 = glm(formula_string, dat4test, family = binomial, na.action = "na.exclude")
        matrix_rsquared[x,y] <-  1 - mod.obj1$deviance/mod.obj1$null.deviance # Pseudo r-squared
        matrix_pvalue[x,y] <- coef(summary(mod.obj1))["predictor",'Pr(>|z|)']
        matrix_tstats[x,y] <- summary( mod.obj1 )$coefficients["predictor","z value"]
      }
    }
  }
  
  rownames(matrix_rsquared) = names(pheno_list)
  rownames(matrix_pvalue) = names(pheno_list)
  colnames(matrix_rsquared) = colnames(data4linear_reg)
  colnames(matrix_pvalue) = colnames(data4linear_reg)
  rownames(matrix_tstats) = names(pheno_list)
  colnames(matrix_tstats) = colnames(data4linear_reg)
  
  matrix_pvalue_df = setNames(reshape2::melt(matrix_pvalue), c("phenotype","module","nom_p"))
  matrix_rsquared_df = setNames(reshape2::melt(matrix_rsquared), c("phenotype","module","rsquared"))
  matrix_tstats_df = setNames(reshape2::melt(matrix_tstats), c("phenotype","module","tstats"))
  all_stats_df = matrix_pvalue_df %>% left_join(matrix_rsquared_df) %>% left_join(matrix_tstats_df) %>%  arrange(nom_p)

  return(list(all_stats_df = all_stats_df, matrix_rsquared = matrix_rsquared, matrix_pvalue = matrix_pvalue, matrix_tstats = matrix_tstats_df))
}

plot_module_trait_association_heatmap <- function(res_test, to_show, show_only_significant = F, signif_cutoff = c("***","**","*")){
  library(ComplexHeatmap)
  library(circlize)
  library(RColorBrewer)

  matrix_rsquared = res_test$matrix_rsquared
  matrix_pvalue = res_test$matrix_pvalue
  
  matrix_rsquared_to_plot = matrix_rsquared[,to_show]
  matrix_pvalue_to_plot = matrix_pvalue[,to_show]
  
  # Adjust P-values by each phenotype separately.
  adj_matrix_pvalue_to_plot = matrix_pvalue_to_plot
  for(i in 1:nrow(matrix_pvalue_to_plot)){
    adj_matrix_pvalue_to_plot[i,] = p.adjust(matrix_pvalue_to_plot[i,], method = "bonferroni")
  }
  adj_matrix_pvalue_to_plot.signif <- symnum(adj_matrix_pvalue_to_plot, corr = FALSE, na = FALSE, 
                                             cutpoints = c(0, 0.001, 0.01, 0.05, 0.1, 1), 
                                             symbols = c("***", "**", "*", ".", " "))
      
  log_matrix_pvalue_to_plot = -log10(matrix_pvalue_to_plot)
  dimnames(log_matrix_pvalue_to_plot) = dimnames(log_matrix_pvalue_to_plot)
  
  if(show_only_significant){
    if(is.numeric(signif_cutoff)){
      to_keep = colSums(adj_matrix_pvalue_to_plot <= signif_cutoff) > 0
    }else{
      to_keep = rep(F,ncol(adj_matrix_pvalue_to_plot.signif))
      for(cut_i in signif_cutoff){
        to_keep = to_keep | colSums(adj_matrix_pvalue_to_plot.signif == cut_i) > 0 # change for the significance you want 
      }
    }
    log_matrix_pvalue_to_plot = log_matrix_pvalue_to_plot[,to_keep,drop=F]
    adj_matrix_pvalue_to_plot.signif = adj_matrix_pvalue_to_plot.signif[,to_keep,drop=F]
  }
  
  matrix_pvalue_to_plot_labels = formatC(log_matrix_pvalue_to_plot, format = "f", digits = 2)
  log_matrix_pvalue_to_plot_t = t(log_matrix_pvalue_to_plot)
  
  # Colored by -log10(pvalue)
  # Numbers inside cell = -log10(pvalue): nominal
  Heatmap(log_matrix_pvalue_to_plot_t, name = "-log10(P-value)",
          cell_fun = function(j, i, x, y, width, height, fill) {
            if(as.character(t(adj_matrix_pvalue_to_plot.signif)[i,j]) == " "){
              grid.text( t(matrix_pvalue_to_plot_labels)[i,j], x, y, 
                        gp = gpar(fontsize = 8))
            }else{
              grid.text(paste0( t(matrix_pvalue_to_plot_labels)[i,j],"\n", t(adj_matrix_pvalue_to_plot.signif)[i,j] ), x, y, 
                        gp = gpar(fontsize = 8))
            }
          },
          col = colorRampPalette(rev(brewer.pal(n = 7, name ="RdYlBu")))(100),
          row_names_side = "left", show_row_names = T,
          cluster_rows = F, cluster_columns = F,
          column_names_gp = gpar(fontsize = 9),
          row_names_gp = gpar(fontsize = 9),
          show_row_dend = F, show_column_dend = F, rect_gp = gpar(col = "white", lwd = 1))
}

work_dir = "/pastel/Github_scripts/SpeakEasy_dlpfc/sn_dlpfc/2nd_pass/eigen_reports/save_lr_adjcov/"
se_results_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/results/"
source("/pastel/Github_scripts/SpeakEasy_net/downstream_speakeasy/speakeasy_downstream_lib.R") # Our Fisher test script. Pvalue adjusted by Bonferroni 

load("/pastel/projects/spatial_t/pseudo_bulk/phenotypes.RData") # phenotypes
ids_dir = "/pastel/resources/ids_datasets/"
```

```{r echo=TRUE}

mypal = c("#1f77b4", # ast
          "#ff7f0e", # end
          "#2ca02c", # ext
          "#d62728", # inh
          "#9467bd", # mic
          "#8c564b", # opc
          "#e377c2"  # oli
)
mypal = setNames(mypal,c("ast","end","ext","inh","mic","opc","oli"))

cell_names = c("Astrocytes", "Endothelial cells", "Excitatory Neurons", "Inhibitory Neurons", "Microglia", "Oligodendrocytes", "OPCs")
cell_names = setNames(cell_names, c("ast","end","ext","inh","mic","oli","opc"))

pheno_list = c("cogng_demog_slope"="gaussian", # Cognitive decline slope. Remove the effect of demog 
               "cogng_path_slope"="gaussian", # Resilience, removed the effect of path + demog
               "tangles_sqrt"="gaussian", # Tangle density - Mean of 8 brain regions
               "amyloid_sqrt"="gaussian", # Overall amyloid level - Mean of 8 brain regions
               "gpath"="gaussian", # Global burden of AD pathology based on 5 regions
               "tdp_cs_6reg"="gaussian", # TDP-43, 6 region severity summary
               "ad_dementia_status"="binomial" # Clinical AD # CT = MCI + NCI 
               )

```

# Load sn
```{r}
sn_columbia <- read.table(paste0(ids_dir, "projid_sn_columbia.txt"), header = T)
sn_columbia$projid <- sprintf("%08d", sn_columbia$projid)

load(paste0(work_dir,"/MIT_pseudobulk_norm.RData"))
```

# Read sn modules 

```{r}
se_results_dir = "/pastel/projects/speakeasy_dlpfc/SpeakEasy_singlenuclei/2nd_pass/snakemake-sn/results/"
# Load sn modules
sn_modules = data.frame()
for(cell_type in c("ext","inh","ast","oli","mic","end","opc")){
  # Filter modules with lest than 30 genes
  modules_file = read.table(paste0(se_results_dir,cell_type,"/geneBycluster.txt"), header = T)
  modules_file = modules_file[, c("ensembl", "gene_name", "cluster_lv3")]
  modules_file$module_clusters = paste0(cell_type,"_M", modules_file$cluster_lv3)
  modules_file$cluster_lv3 = NULL
  
  modules_file = modules_file %>% dplyr::group_by(module_clusters) %>%
    dplyr::mutate(module_size = n()) %>% dplyr::filter(module_size >= 30) %>%
    dplyr::mutate(cell_type = cell_type)
  
  sn_modules = rbind(sn_modules, modules_file)
}
# length(unique(sn_modules$module_clusters)) # 193
# modules by cell type
table(unique(sn_modules[,c("module_clusters","cell_type")])$cell_type) # 7
```

# Project sn modules in the MIT data

We already have the modules from the Columbia sn-ROSMAP, so I'll use them as a gene list and calculate the eigengenes and module average expression for the MIT sn-ROSMAP dataset. Then, we’ll run regressions with AD-related covariates to check for associations.

Heatmaps: Bonferroni adjusted by column. 

## Astrocytes 
We keep only the MIT samples NOT in Columbia sn. 
```{r}
# Astrocytes 
net_output = unique(sn_modules[sn_modules$cell_type == "ast", ])
expr_mtx_sel = as.data.frame(celltype_exp$ast$tmm_voom)[, ! colnames(celltype_exp$ast$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_ast.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_ast.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_ast.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=8}
res_test = run_module_trait_association(data4linear_reg = data4linear_reg, 
                                        phenotype_dt = phenotype_dt, 
                                        pheno_list = pheno_list, 
                                        covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_ast.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=4, fig.width=5}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## Microglia 
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "mic", ])
expr_mtx_sel = celltype_exp$mic$tmm_voom[, ! colnames(celltype_exp$mic$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_mic.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_mic.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_mic.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=10}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_mic.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At lemic one module with adjusted pvalue < 0.05. 

```{r fig.height=3.5, fig.width=4}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## Excitatory neurons
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "ext", ])
expr_mtx_sel = celltype_exp$ext$tmm_voom[, ! colnames(celltype_exp$ext$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_ext.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_ext.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_ext.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=9}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_ext.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=4, fig.width=4}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## Inhibitory neurons
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "inh", ])
expr_mtx_sel = celltype_exp$inh$tmm_voom[, ! colnames(celltype_exp$inh$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_inh.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_inh.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_inh.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=8}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_inh.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=3.5, fig.width=4.5}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## Oligodendrocytes 
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "oli", ])
expr_mtx_sel = celltype_exp$oli$tmm_voom[, ! colnames(celltype_exp$oli$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_oli.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_oli.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_oli.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=10}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_oli.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=6, fig.width=5}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## Endothelial cells 
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "end", ])
expr_mtx_sel = celltype_exp$end$tmm_voom[, ! colnames(celltype_exp$end$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_end.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_end.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_end.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=8}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_end.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=2.5, fig.width=4.5}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

## OPCs
```{r}
net_output = unique(sn_modules[sn_modules$cell_type == "opc", ])
expr_mtx_sel = celltype_exp$opc$tmm_voom[, ! colnames(celltype_exp$opc$tmm_voom) %in% sn_columbia$projid]
rownames(expr_mtx_sel) <- gsub("(*.)\\.(.*)","\\1", rownames(expr_mtx_sel))
expr_mtx_sel$ensembl = rownames(expr_mtx_sel)
gene_mod_map = data.frame(ensembl = rownames(expr_mtx_sel))
gene_mod_map = gene_mod_map %>% inner_join(unique(net_output[,c("ensembl","module_clusters")]), by = c("ensembl"))
gene_mod_map$cluster_lv3 = gsub("(.*)_M(.*)","\\2",gene_mod_map$module_clusters)

expr_mtx_sel_ord = expr_mtx_sel[match(gene_mod_map$ensembl,rownames(expr_mtx_sel)),] # order the matrix
identical(rownames(expr_mtx_sel_ord), gene_mod_map$ensembl) # must be TRUE 

colors_mod = gene_mod_map$cluster_lv3
expr_mtx_sel_ord$ensembl = NULL
expr_mtx_sel_ord_t <- data.frame(t(as.matrix(expr_mtx_sel_ord)))
external_ME = moduleEigengenes(expr_mtx_sel_ord_t, colors = colors_mod)
mod_average = external_ME$averageExpr # data.frame

# Save results 
save(external_ME, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_opc.Rdata"))
write.table(external_ME$eigengenes, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleEigengenes_lv3_opc.txt"), sep = "\t", quote = F, row.names = T)
write.table(external_ME$averageExpr, file = paste0(work_dir, "replication_analysis_filt/MIT_moduleaverageExpr_lv3_opc.txt"), sep = "\t", quote = F, row.names = T)
```

### Regressions 
```{r message=TRUE}
data4linear_reg <- mod_average # match codes name 
phenotype_dt = phenotypes[match(rownames(data4linear_reg), phenotypes$projid), ]

all(rownames(data4linear_reg) == phenotype_dt$projid) # Must be TRUE. Match IDs
# hist(phenotype_dt$amyloid_sqrt)
message(paste0("Unique participants: ", length(unique(phenotype_dt$projid)))) 
```

```{r, message=FALSE, results='hide', warning=FALSE, fig.width=4, fig.height=8}
res_test = run_module_trait_association(data4linear_reg, phenotype_dt, pheno_list, covariates = c("age_death","msex", "educ")) # covariates to adjust
matrix_rsquared = res_test$matrix_rsquared
matrix_pvalue = res_test$matrix_pvalue
save(res_test, file = paste0(work_dir, "replication_analysis_filt/MIT_res_test_opc.Rdata"))

plot_module_trait_association_heatmap(res_test)
```

### Top results
Top result by covariate. 

```{r}
# res_test$all_stats_df 
createDT(res_test$all_stats_df %>% group_by(phenotype) %>% slice_head(n = 1))
```

### Nominal pvalue
```{r}
createDT(matrix_pvalue)
```

### Significant results
Threshold: At least one module with adjusted pvalue < 0.05. 

```{r fig.height=3, fig.width=5}
plot_module_trait_association_heatmap(res_test, show_only_significant = T, signif_cutoff = 0.05)
```

# Session info
```{r}
sessionInfo()
```
